# Introduction {#sec-intro}

There are inherent limitations of clinical trials
(RCTs) done before a medication is authorised in terms of cost,
surveillance time, size and a lack of diversity of patients included
(e.g. that pregnant women, children and elderly often are not included)
[@sanson2007limitations]. Another issue is that RCTs are limited in
their ability to detect rare adverse drug reactions (ADRs) and potential
drug interactions (DDIs) since they usually only include a few thousand
participants at most. Moreover, the concurrent use of multiple medications is often
an exclusion criterion in RCTs, leaving gaps in the assessment of
polypharmacy risks. As a result, tested drug combinations tend to be
limited to those where physicians and pharmacologists have existing
clinical experience or hypotheses about potential interactions
[@van2002assessment]. This means that monitoring of the safety of
medications and the potential for DDIs after marketing is essential.
This monitoring is often referred to as post-authorization
pharmacovigilance.

Elderly individuals are commonly prescribed multiple medications due to
the increasing prevalence of comorbidities associated with aging. One
study suggested that patients aged $75$ or older in Austria took on
average $7.5$ drugs [@schuler2008polypharmacy]. Among this demographic,
$10\%$ of hospitalizations were attributed to ADRs, and $18.7\%$ of
these ADRs were caused by a combination of more than one drug, a
drug-drug interaction. More broadly, a meta-analysis attributed
approximately $22.2\%$ of hospitalizations caused by adverse drug
reactions to DDIs [@dechanont2014hospital].

Disproportionality analysis (DA) encompasses widely used methods in
pharmacovigilance to detect ADRs by assessing the frequency of adverse
events (AE) associated with specific drug consumption relative to what would
be expected. Most established disproportionality methods like the
proportional reporting ratio (PRR) [@evans2001use], reporting odds ratio
(ROR) [@vanPuijenbroek2002ACO], Bayesian Confidence Propagation Neural
Network (BCPNN) [@bate1998bayesian], and the Gamma Poisson Shrinker
(GPS) [@dumouchel1999bayesian] allow for the identification of signals
that warrant further investigation in the context of single drug
assessment. The tree-based scan statistic [@kulldorff2003tree;@heo2024signal] takes advantage of the tree structure of the Anatomical Therapeutic Chemical (ATC) drug classification in order to allow to select a drug family rather than a single drug, a family corresponding to a subtree of the ATC tree. Likelihood-ratios tests then allow to choose the most relevant family.

In order not to overinterpret high proportions in case of rare drug cocktails, it is possible to derive a score from an independence test on the contingency table counting the presence or absence of the adverse event depending on the consumption of the drug cocktail. [@gosho2017utilization] use a chi-square statistic to do so, while [@ahmed2010false] rather use the Fisher exact test which has the advantage of being non-asymptotic.

Fewer DA methods are available to use for multiple drug
consumption, such as the $\Omega$ shrinkage method
[@noren2008statistical]. This method allows the use of a DA measure in
order to detect sets of the type *Drug-Drug-Adverse Event* and stands as
the standard measure when it comes to detecting the interactions of two
drugs. Adaptations of the PRR for a single drug has been proposed like the Concomitant Signal Score (CSS) [@noguchi2020improved] and the PRR adaptation for DDIs [@wang2020propensity].


Other methods than DA measures exist in order to detect DDIs. Among
them, well known methods encompass logistic models [@van2000detecting;
@van1999signalling] and association rules methods [@noguchi2018new;
@ibrahim2016mining]. One can refer to multiple reviews for a more
detailed overview of existing methods [@ibrahim2021signal;
@hauben2023artificial]. 

Computational pharmacovigilance methods are typically applied to Spontaneous Reporting Systems (SRS), which aggregate Individual Case Safety Reports (ICSRs)—real-world data submitted by healthcare professionals, manufacturers, and patients. Such reports contain information about intake of medications for each patient as well as their experienced AE. While SRS databases, such as the FDA Adverse Event Reporting System (FAERS), provide the scale necessary to detect rare events, they possess inherent limitations. These include under-reporting, where only a fraction of adverse events are captured [@bate2009quantitative], and various reporting biases, such as the notoriety bias [@pariente2007impact]. Crucially, the spontaneity of these reports often results in an incomplete recording of all drugs taken by a patient. Furthermore, because the exact timing of drug administration is frequently omitted, it is difficult to distinguish between simultaneous intake and medications taken at different times within the same reporting window. This temporal ambiguity potentially limits the precision of interaction analyses, as the observed cocktail may represent a sequence of exposures rather than a concurrent pharmacological event. Consequently, because the total number of patients exposed to a drug is unknown, these databases cannot provide true incidence rates, necessitating the use of disproportionality analysis to identify safety signals [@almenoff2005perspectives].

\medskip

While previous work showed that further
practical experiences should be of interest for high-order drug
interaction testing [@Tekin2018prevalence], it is hard to know which
exact combination one should test and easy to miss an important
combination in the overwhelming space of high-order drug cocktails. The
present article proposes a novel method to identify drug cocktails that
are overrepresented in adverse event reports, which constitute candidates
for further interaction assessment. It requires the choice of a scoring function able to measure, for a given AE, the importance of the disproportionality for any medication set, including single medications. It then takes advantage of the tree structure of the ATC classification to explore the space of medication sets in two ways: a genetic algorithm looking for the sets of high score, and a MCMC algorithm learning the distribution of the scores for a given cocktail size in order to evaluate how extreme a high score is. A post-treatment step using penalized logistic regression then allows to assess whether an identified cocktail reflects a true pharmacological interaction or merely the combined effect of its components. The approach is also original with respect to the fact that a medication denotes either a drug or a drug family in the sense of an internal node of the ATC classification.

The developed algorithms, available in the corresponding R package **emcAdr** [@R-emcAdr], aim to guide pharmacology researchers toward drug combinations that might correspond to drug interactions and require more rigorous monitoring.

# Methods

## ATC Tree and Cocktail Definition {#sec-cocktail}

Multiple classifications of drug active ingredients exist, one being the
ATC classification. This system is structured hierarchically and can be
represented as a tree with five levels containing, at the time of this study, $6809$
nodes. The leaves of this tree are the active ingredients, while the
first level consists of nodes representing organs or systems affected by
the descendant active ingredients. The remaining nodes correspond to
therapeutic or pharmacological families.

By applying a Depth-First Search algorithm to enumerate each node of the
ATC tree, we can define a drug cocktail as a set of integers
corresponding to specific nodes in the tree. More formally, a cocktail
$C$ of size $k\geq 1$ is defined by $C = (x_1, \dots, x_k) \in \Delta^{k}$
where $k \in \mathbb{N}$ , and $\Delta$ represents the set of numbered
nodes of the ATC tree $T$. @fig-mut shows examples in a simplified tree,
the green nodes denoting the considered cocktail. Note that for convenience, we also refer to single drugs as cocktails.

Considered cocktails can include internal nodes of the tree (thus
representing families of active ingredients), allowing for the detection
of more general signals. For example, paracetamol might send a weak
signal, while if we move up the tree, analgesics as a whole could
represent a stronger and more general signal. Therefore, all patients
taking at least one drug from this drug family will be considered in the
computation of the risk. That notion is equivalent to the notion of cut in the tree-scan method [@kulldorff2003tree].



## Cocktail Risk Characterization {#sec-risk}

In the field of pharmacology, accurately characterizing the risk associated with drug administration is a complex task. The aim of the developed method is to search the space of cocktails to maximize a score indicating if individuals taking a given drug combination have a higher risk of an AE. While the algorithm is designed to be flexible and can be run with any scoring function, we focus on a comparison of established metrics to justify our selection (see @sec-score_comp).

### Scoring Functions in Pharmacovigilance{#sec-risk-definition}

To evaluate the disproportionality of drug cocktails presence for a particular AE, several statistical measures have been proposed in the literature:

Proportional Reporting Ratio (RR)
: A standard measure defined as the ratio of the probability of an AE occurring in the group exposed to the cocktail versus the non-exposed group. Its application in signal generation from spontaneous reports was notably discussed by [@evans2001use].

PRR for Drug-Drug Interactions (PRR)
: Unlike the single-drug PRR, this adaptation for drug combinations evaluates whether a combination represents a synergistic risk. Specifically, as discussed by [@wang2020propensity], a signal is often defined by comparing the lower bound of the confidence interval of the combination's risk to the maximum risk observed for the individual drug components.

$\Omega$ Shrinkage
: A Bayesian measure specifically developed for drug-drug-event combinations. As proposed by [@noren2008statistical], it utilizes a "shrinkage" approach to reduce false positive signals in cases with limited data by pulling the observed association toward the value expected under independence.

Concomitant Signal Score (CSS)
: Introduced as an improved detection criterion, this metric aims to enhance the detection of DDI signals by building upon the PRR framework, as detailed by [@noguchi2020improved].

Hypergeometric Disproportionality Score
: While the methodology can accommodate various scores, we propose the use of a score derived from the Fisher exact test. As highlighted by [@ahmed2010false], this approach has the advantage of being non-asymptotic. Consider a dataset of $N$ patients, among which $K$ experience the adverse event AE. Let $n_C$ be the number of people taking a cocktail $C$ and $x$ the number of patients taking $C$ and experiencing AE. We define the risk to be:

$$H(C) = -\log(\mathbb{P}(X \ge x))$$

where $X \sim \mathcal{H}(n_C, K, N)$. This measure effectively functions as the log p-value under the null hypothesis that the number of people with AE in a uniform sample of $n_C$ out of $N$ people follows the hypergeometric distribution. It has the distinct advantage of accounting for the total number of patients exposed to the cocktail; for instance, $H(C)$ will assign a higher risk to a cocktail taken by 100 patients with 50 AE cases than to a cocktail taken by 10 patients with 5 AE cases, whereas simple ratios like the RR might treat them identically. Such hypergeometric measures are well-established in other high-dimensional fields like bioinformatics for functional enrichment analysis [@grossmann2007improved].

<!-- In the field of pharmacology, accurately characterizing the risk
associated with drug administration is a hard task. Highly used methods
to evaluate drug interactions are limited to the evaluation of two
concomitant drugs [@noren2008statistical; @wang2020propensity], while
other, like the PRR [@evans2001use; @wang2020propensity], are biased
toward drugs or drugs cocktails taken by few patients, assigning high
risk to them (also see @fig-comparison_score). In order to mitigate this
phenomenon, we propose to use the hypergeometric distribution to
evaluate the hazard of medication cocktails. Consider a dataset of $N$
patients, among which $K$ experience the adverse event AE. Let $n_C$ be
the number of people taking a cocktail $C$ and $x$ the number of
patients taking $C$ and experiencing AE . We define the risk to be $$
H(C) = -\log(\mathbb{P}(X \ge x))
$$ where $X \sim \mathcal{H}(n_C, K, N)$. This measure is the log
p-value under the null hypothesis that the number of people with AE in a
uniform sample of $n_C$ out of $N$ people follow the hypergeometric
distribution. It has the advantage to take into account the relative
number of patients taking the cocktail $C$. For example, $H(C)$ will be
greater for a cocktail taken by $100$ patients, among whom $50$
experience AE, than for a cocktail taken by 10 patients, among whom 5
experience AE. In contrast, the PRR @evans2001use makes no distinction
between the two.

Such measures are commonplace in bioinformatics, especially when working
on functional enrichment analysis @grossmann2007improved. -->

## Distinguishing Combined Risk from True Interaction{#sec-firth}

While the $H(C)$ score effectively identifies cocktails associated with a significant increase in the AE frequency, it does not inherently distinguish a synergistic pharmacological interaction from situations corresponding to the independent addition of individual effects or an effect driven by a subset of the cocktail (e.g., "innocent" medications co-prescribed with a high-risk drug).

To address this, we propose a post-treatment of detected signals using Firth’s penalized likelihood logistic regression [@firth1993bias]. This method is particularly suited for spontaneous reporting systems as it reduces bias in maximum likelihood estimates and provides a solution to the problem of separation, which often occurs with rare adverse events. Using the *logistf *R package [@R-logistf], we estimate a model for a given cocktail $C$ (e.g., $d_1, d_2$) as follows:

$$\text{logit}(P(AE=1)) = \beta_0 + \beta_1 \mathbb{1}_{d_1} + \beta_2 \mathbb{1}_{d_2} + \beta_{3} \mathbb{1}_{d_1 \cap d_2}$$

An interaction is confirmed if the interaction coefficient $\beta_{3}$ is positive and statistically significant. This indicates that the risk associated with the co-prescription exceeds the additive effect of the individual components on the logit scale. While the example above illustrates the case of a size-two cocktail for simplicity, this post-treatment generalizes naturally to cocktails of any size $k$ by including all $2^k - 1$ interaction terms in the model, or by testing nested models comparing the full cocktail to its sub-cocktails. Users may implement the post-treatment of their choice. 

## Disproportionality Identification

### High-risk Drug Cocktails Identification {#sec-genetic}

The number of possible cocktails is $2^L$, where $L$ denotes the number of nodes in the ATC tree, and even for a given cocktail size $k$, the number of possibilities is still ${L}\choose{k}$. It is
therefore not possible to compute $H(C)$ for each possible cocktail $C$
included in the dataset. Instead, to explore the space of cocktails and
locate those at high-risk of AE, we use a genetic algorithm
[@petrowski2017evolutionary]. It simulates the evolution of a population of cocktails according to the principle of natural selection, in order to search for the most performing individuals with respect
to an evaluation criterion based on $H$. 

The steps are
the following, the algorithm repeating steps 2 and 3 until a user-defined number
of iterations is reached.

Initialization.

:   The genetic algorithm's population consists of $m$ cocktails. These
    cocktails are randomly initialized and can vary in size.

Evaluation and selection.

:   At iteration $n$, the current population $P_n$ undergoes an evaluation and selection phase. The evaluation computes the score $H(C)$ for each cocktail $C$ in the population.  A new population $Q_{n+1}$ of the same size is drawn by sampling $m$ times a pair of cocktails in the original population and copying the one with highest score. Note that this step performs selection as the expectation of the number of copies in $Q_{n+1}$ of a cocktail from $P_n$ is an increasing function of its score. A penalization is however applied to avoid a uniformization of the population,  as further explained in @sec-diversity.

Stochastic modification.

:   Mimicking the genetic drift in nature, stochastic modifications occur in the population in order to explore the large space of cocktails. First, a
    crossover operation allows two cocktails to exchange information.
    Here, the crossover involves exchanging sub-trees between two
    cocktails as follows:

:   -   an internal node $v$ of the ATC tree is randomly selected;

:   -   the nodes of the subtree rooted at $v$ are exchanged between the
        two sequences.

:   After performing the crossover, a mutation is applied to the
    resulting individuals, chosen from two types. The first type is a
    local mutation which changes a randomly selected node of the
    cocktail to one of its free neighboring nodes. This mutation is
    further explained in @sec-MCMC. The second type is an
    addition/deletion mutation which operates as follows, with $k$ being
    the cocktail length and $\alpha$ a chosen hyperparameter:

:   -   with probability $\min(1,\frac{\alpha}{k})$, a node uniformly
        drawn from $\Delta$ is added to the sequence;

:   -   with probability $1 - \min(1,\frac{\alpha}{k})$, a uniformly
        drawn node from the cocktail is removed.

:   An example of crossover and mutations can be seen on @fig-mut
    (a), (c) and (d).
    
:   Applying those modifications to $Q_{n+1}$ yields a population $P_{n+1}$ which is used to loop at step $n + 1$.

![Cocktail modifications used for the genetic algorithm (a, c and d) and
the MCMC algorithm (b and c). Green nodes are part of considered
cocktails. In Crossover, the orange node represents the selected
internal node whose subtrees are being swapped. In local mutations,
orange nodes represent legal moves](figures/all_mutations.svg){#fig-mut}

### Distance Between Drug Cocktails and Cocktail Penalization {#sec-diversity}

If at some point of the algorithm, an individual of $P_n$ has a high score compared to other individuals,  the genetic algorithm may converge to a
population consisting entirely of that cocktail. To avoid this
uniformization phenomenon, similar cocktails are penalized in the
evaluation phase as follows,

$$H_{pen}(C) = \frac{H(C)}{\sum_{C_i \in \mathcal{C}} \text{Sim}(C,C_i) }$$

The computation of the similarity $\text{Sim}(C,C')$ is based on a
distance inspired by the Levenshtein distance [@levenshtein1966binary].
However, unlike the traditional Levenshtein distance, sequences are
treated as unordered sets. For two drug cocktails, $C_1$ and $C_2$, of
sizes $n_1$ and $n_2$, the distance $d(C_1 , C_2)$ is defined as the
minimal cost required to transform $C_1$ into $C_2$ using three
elementary operations.

-   $\text{Ins}_a(C)$ consists of adding $a$ to the cocktail $C$.
-   $\text{Del}_a(C)$ consists of deleting $a$ from the cocktail $C$.
-   $\text{Sub}_{a,b}(C)$ consists of substituting $a \in C$ by $b$.

An associated cost is defined for each operation.

Substitution.

:   The cost associated with the substitution operation is chosen to be
    consistent with the conceptual similarity of cocktails. If $a$ is a
    drug belonging to $C$, the cost should increase as the drug $b$
    diverges further from drug $a$. For example, if $a$ is a drug, and
    $b$ a drug family that contains $a$, the cost should be moderate.
    Conversely, if $b$ is a drug family not containing $a$ the cost
    should be higher. This distance is thus defined by the maximal
    distance between $a$ and $b$ to their Lowest Common Ancestor.

Insertion, Deletion.

:   The deletion and insertion cost are chosen as $\frac{depth(T)}{2}$.
    This choice implies that a substitution always costs less than a
    deletion followed by an insertion. The latter are used only when the
    two cocktails do not have the same length.

A transformation $f$ from $C_1$ to $C_2$ is a composition of elementary
operations that go from $C_1$ to $C_2$. The associated cost
$\text{cost}(f)$ is defined as the sum of the cost of the operations
used in $f$. Finally,
$d(C_1,C_2) = \min\limits_{f} \text{cost}(f : f(C_1) = C_2 )$.

Finally, the maximum distance between $C_1$ and $C_2$ being
$(n_1 + n_2) \frac{depth(T)}{2}$, we define the similarity as
$$ Sim(C_1,C_2) = 1 - \frac{2D(C_1,C_2)}{(n_1 + n_2)depth(T)}$$

The computation of the similarity is achievable in
$\mathcal{O}(n_1 \times n_2 \times   \text{depth}(T) + |\Delta|)$ operations
in the worst case. The algorithm to compute the similarity
is detailed in the @sec-appA (see @algo-distance).

### Output Clustering {#sec-clust}

Despite the diversity mechanisms integrated into the genetic algorithm,
not all drug cocktails retrieved within the genetic algorithm population
are unique. It is moreover common to encounter solutions that are merely
variations of others, differing only by transformations such as changing
a node in the ATC tree to its parent or child. To streamline analysis
and enhance the efficiency, a post-treatment clustering of similar drug
cocktails is implemented. This method allows to focus on the most risky
cocktails within each cluster or to interpret pharmaceutically clusters
rather than individual cocktails.

To do so, drug cocktails are embedded into a two-dimensional space using
the UMAP algorithm [@mcinnes2018umap], which aims to preserve similarity
in the latent space. This representation enables the effective use of
conventional machine learning clustering algorithms in $\mathbb{R}^2$.
Specifically, the DBSCAN algorithm [@ester1996density] is applied to
identify clusters of similar drug cocktails with an intuitive way of
choosing hyperparameters.

## Approximate P-Value Assignment for Drug Cocktails {#sec-MCMC}

Once a list of cocktails with high score has been established by the genetic algorithm, an important step of the analysis is to assign them p-values to decide if they are significant.

Consider a cocktail $C$ of size $k$ in the list and its score $S_{obs}$.  Denote by $H_0$ the null hypothesis according to which a cocktail does not favor the AE, and by $N_0$ the number of cocktails of size $k$ for which $H_0$ is the truth. Similarly, $H_1$ represents the alternative hypothesis of a cocktail favoring the AE and $N_1$ is the number of such cocktails of size $k$. The p-value corresponding to $S_{obs}$ is then  $\mathbb{P}_{H_0}(S>S_{obs})$, but cannot be estimated without a modelling hypothesis under $H_0$.

However, if $\mathbb{P}_M (S)$ refers to the marginal distribution, that is the score distribution for the cocktails both in $H_0$ and $H_1$,

$$\mathbb{P}_M(S>S_{obs})  = \mathbb{P}_{H_0}(S>S_{obs}) \frac{N_0}{N_0+N_1} + \mathbb{P}_{H_1}(S>S_{obs}) \frac{N_1}{N_0+N_1}$$
so that 

$$\frac{N_0+N_1}{N_0}\mathbb{P}_M(S>S_{obs}) - \frac{N_1}{N_0} \mathbb{P}_{H_1}(S>S_{obs}) \leq  \mathbb{P}_{H_0}(S>S_{obs}) \leq \frac{N_0+N_1}{N_0}\mathbb{P}_M(S>S_{obs})$$
Under the reasonable assumption $N_1 << N_0$, the probability $\mathbb{P}_M(S>S_{obs})$ on the marginal distribution can thus be used as an approximation of the p-value. Note that  the upper bound yields that, under the weak hypothesis that less than $10\%$ of the cocktails of size $k$ favor the AE, the real p-value is at most equal to $1.11$ times its approximation. Moreover, the approximated p-value can be estimated by computing the score of cocktails sampled in the general cocktail population. The proposed method therefore focuses on a sampling scheme in the general population and uses the approximated p-value to declare significance.

A naive sampling of cocktails considers almost only cocktails taken by no patient in the dataset.
A Metropolis-Hastings MCMC algorithm is thus considered, as it can be used by conditioning on the fact that all visited cocktails are present in the dataset [@au2001estimation].

To employ such an algorithm, it is necessary to define a state space
$\mathcal{C} = \{C_{1}, \dots, C_{p}\}$, a computable target measure
$f(C_{i})$, and conditional laws $q(.|C_{i})$ under which simulation is
possible and new states can be proposed.

State set.

:   The state set is made of all cocktails of $k$ drugs for a fixed $k$.

Proposal law.

:   The proposal law is defined as a mixture of two mutation laws of the
    current cocktail. They operate as follows:

:   -   Random mutation consists of a completely random movement in the
        cocktail space.

:   -   Local mutation involves a local movement relative to the
        structure of the drug tree. Here, a node $x_p$ of the state
        $C_i$ is changed to one of its free neighboring nodes.

At each iteration, the random and local mutations have probability $p_R$
and $1-p_R$, where $p_R$ is a hyperparameter.

@fig-mut (b, c) presents examples of a random and a local mutation.

State evaluation.

:   The evaluation of a drug cocktail is based on the score $H(C)$. The
    chosen target measure is then:
    $$f_{T}(C_i) = \frac{1}{Z(T)} \times {e^{\frac{H(C_i)}{T}}}$$

:   where $Z(T) = \sum_{C} e^{\frac{H(C)}{T}}$. $T$ is a parameter known
    as temperature, which modulates space exploration by more readily
    accepting cocktails with moderate scores (high $T$) or, conversely,
    by strongly favoring combinations of drugs with high scores (low
    $T$).

:   The acceptance probability of cocktail $C_{i+1}$ from cocktail
    $C_{i}$ is given by:

$$\min\left(1,\frac{f_T(C_{i+1})}{f_T(C_{i})} \times \frac{q(C_{i}|C_{i+1})}{q(C_{i+1}|C_{i})}\right)$$

The theory related to the Metropolis-Hastings algorithm [@Robert2004]
ensures that the empirical distribution of $f_T(C_i)$ for the
constructed cocktail chain converges to the distribution of $f_T(C)$. A
very long realization of such a walk, therefore, allows for the
approximation of the distribution that can be used to determine an
empirical p-value for the score of a cocktail of interest. This enables
the possibility to say whether or not a high-risk is truly significant
(e.g. among the top 5% of scores). It defines what a high-risk is in our
case.

## Datasets {#sec-datasets}

### Simulated Data {#sec-simu_data}

Multiple datasets were simulated to evaluate the method performance
against known outcomes. The datasets, designed to challenge the
algorithm, simulate various patient scenarios. Each patient record
includes prescribed medications and the corresponding occurrence of an
adverse event $AE$.

The first dataset ($D_1$) is composed of 200,000 patients, and has the following
characteristics:

-   $1\%$ of the patients take a size-3 drug cocktail $C_1$ and have a
    $\frac{1}{100}$ chance of having $AE$.
-   $1\%$ take a size-3 drug cocktail $C_2$ and have a $\frac{1}{200}$
    chance of having $AE$.
-   $1\%$ take a size-2 drug cocktail $C_3$ and have a $\frac{1}{100}$
    chance of having $AE$.
-   $1\%$ take a size-2 drug cocktail $C_4$ and have a $\frac{1}{200}$
    chance of having $AE$.

A small percentage of the dataset ($1.5 \%$ per combination) are
combinations of 2 out of the 3 drugs from $C_1$ and $C_2$, but with no
risk of $AE$. This helps to mitigate the false identification of
sub-cocktails of $C_1$ and $C_2$ as high-risk cocktail because those who
take two drugs of these cocktails will almost surely take the remaining
drug of the cocktail.

The remaining $87\%$ of the dataset consists of patients assigned
random cocktails drawn uniformly. The size $s$ of each cocktail is drawn
according to a Poisson distribution with $\lambda = 4$ (mean size of
drugs cocktails taken by patients in the dataset). For each cocktail, $s$
nodes of the ATC tree are selected uniformly, with each combination
assigned an adverse event with probability $\frac{1}{15000}$.

All simulations use the complete ATC classification tree comprising 6809 nodes
across five hierarchical levels at the time of this study. The cocktails $C_1$ through $C_4$ inducing an elevated AE probability were chosen as nodes at level 4 of the ATC hierarchy (i.e., pharmacological subgroups, one level above the leaf nodes representing individual active substances). This choice is deliberate: in the simulated data, patients are assigned specific drugs at level 5 (leaves) as would occur in practice, but any patient taking a drug that descends from one of the $C_i$ nodes inherits the elevated AE probability associated with that family. This design tests the method’s ability to detect family-level signals — an important feature, since the underlying pharmacological mechanism may apply to an entire drug class rather than a single substance. The $C_1$ through $C_4$ nodes were selected to be non-overlapping (i.e., belonging to distinct subtrees) and to have a sufficient number of descendant leaves to generate realistic prescription patterns.  

Three additional datasets were similarly constructed, varying the sizes of the
cocktails inducing AE. Dataset $D_1$ contains cocktails of sizes two and three as
described above. Dataset $D_2$ contains only size-two cocktails, Dataset $D_3$ only
size-three cocktails, and Dataset $D_4$ contains size-two, three, and four cocktails.

### FAERS Data {#sec-FAERS}

As a proof-of-concept on real data, the method was also assessed using the FAERS dataset, which consists of
ICSRs submitted by healthcare professionals, consumers, and
manufacturers. These reports include details on patient drug intake and
the side effects experienced. The methods were deployed on FAERS data from the second quarter of $2013$ to the second quarter of $2015$, the restriction to a two-year window being motivated by computational time constraints.

Significant refinement of the FAERS data was required. Duplicate reports
were removed, retaining only the report with the most recent ID.
Subsequently, a link was established between the prescribed drugs and
the ATC codes of each active ingredient. This process involved matching
drug names to their respective active ingredients and converting these
ingredients to their corresponding ATC codes. The DiAna dictionary
[@fusaroli2024enhancing] facilitated the standardization of FAERS drug
names to ATC codes. It is important to note that the DiAna dictionary handles combination products by splitting them into their constituent active substances rather than assigning a single ATC combination code. <!-- Within our framework, this decomposition is advantageous as it allows the methodology to treat combination products as sets of individual ingredients facilitating the detection of risk signals driven by specific chemical components across different formulation --> Reports with unmatchable drug names in the DiAna
dictionary were excluded, reducing the dataset from 2,043,231 to
1,612,931 patients.

For this study, we focused on myopathy as the selected adverse event
outcome. It is a clinically concerning condition with a sufficient
number of reported cases in the dataset ($536$ cases). To validate our
results, we compared the identified drug-myopathy associations with
known drugs already established to cause myopathy [@miernik2024drug;
@hall2011musculoskeletal;@valiyil2010drug]. Code for data refinement is
available [on GitHub](https://github.com/JulesBa-Git/FAERS-xml-parser).

# Results

## Score Comparison{#sec-score_comp}

To support the use of the hypergeometric score $H(C)$ introduced in
@sec-risk, we compared it to other risk scores on Dataset $D_2$, which comprises pairwise drug combinations. A set of cocktails was generated under the alternative hypothesis $H_1$, meaning they correspond to the drug combinations specifically designed to increase the probability of AE in Dataset $D_2$ (see @sec-simu_data). A second set of cocktails was sampled under the null hypothesis $H_0$, representing a baseline risk without specific interaction. For each cocktail in both sets, the risk scores were computed using the patient data in Dataset $D_2$. @fig-comparison_score illustrates the performance of the risk scoring methods for detecting high-risk drug combinations introduced in @sec-risk-definition. Each subplot displays the score values for cocktails representing true solutions, sampled under $H_1$ (green) and cocktails not representing true solutions, sampled under $H_0$ (red). The bottom-right panel presents the Precision-Recall
(PR) curve, comparing the detection power of the scores in identifying
high-risk cocktails. PRR is a special score because its value is either
one or zero, which allows the computation of only one value of precision
and recall. $PRR$ is therefore represented by a single point rather than a
PR curve.

PR curves of the different scoring methods are mainly comparable, confirming the conclusions of [@candore2015comparison]. However, as illustrated by the jitter plots,  the hypergeometric score and the $\Omega$ Shrinkage measure better rank and isolate true solutions from other cocktails. It must be emphasized that the $\Omega$ Shrinkage
measure is difficult to compare using a threshold, as the original
article [@noren2008statistical] suggests signaling a cocktail when the
score exceeds zero. However, in the simulations, the computed score
never exceeded zero, as depicted by the x-axis of its jitter plot.
Despite these considerations, these two methods are the least biased
toward cocktails taken by only a few patients.

The good behavior of the hypergeometric score and its easy generalization to cocktails of any size justify the use of that scoring function for the exploration of the cocktail space.

<!-- ![Comparison of scores for cocktail of size-two. Each dots denotes a cocktail while his risk computed on the formerly presented synthetic dataset (@sec-simu_data) is showed on the x-axis. Red dots represent cocktails that do not induce adverse event (negative controls), green ones represent cocktail inducing adverse event (positive controls). The bottom-right corner shows the Precision-Recall curves for each score. The perfect classification corresponds to the upper right corner. The areas under the curves allow to compare different methods](figures/comparison_metrics_omega_025.svg){#fig-comparison_score} -->


::: {#fig-comparison_score}
```{r}
#| fig-show: hold
#| layout: [[50, 50], [50,50],[-10,80,-10]]
#| warning: false
#| message: false
#| echo: false

library(emcAdr)
library(dplyr)
library(dbscan)
library(ggplot2)
library(umap)
library(stringr)

load("figures/datasets/PRC_without_computation.RData")

Metrics_computation_size2_1M_count$omega_025[Metrics_computation_size2_1M_count$omega_025 <= -2e+09] <- -4
zero_score <- Metrics_computation_size2_1M_count[Metrics_computation_size2_1M_count$RR == 0, ]
non_zero_score <- Metrics_computation_size2_1M_count[Metrics_computation_size2_1M_count$RR != 0, ]
set.seed(42)

zero_score_sample <- zero_score[sample(nrow(zero_score), size = 0.05 * nrow(zero_score)), ]
reduced_data <- rbind(non_zero_score, zero_score_sample)

ggplot(reduced_data, aes(x = RR, y = as.factor(Label), color = as.factor(Label))) +
  geom_jitter(width = 0, height = 0.1, size = 2, alpha = 0.7) +
  scale_color_manual(values = c("red", "limegreen")) +
  labs(
    title = "RR by Label",
    x = "RR",
    y = "Label",
    color = "Label"
  ) +
  theme_minimal(base_size = 16) + 
  theme(
    plot.title = element_text(size = 20, hjust = 0.5), 
    axis.title = element_text(size = 16),              
    axis.text = element_text(size = 14),                              
    legend.title = element_text(size = 16),            
    legend.text = element_text(size = 14),                            
    legend.position = "right"
  )

ggplot(reduced_data, aes(x = phyper, y = as.factor(Label), color = as.factor(Label))) +
  geom_jitter(width = 0, height = 0.1, size = 2, alpha = 0.7) +
  scale_color_manual(values = c("red", "limegreen")) +
  labs(
    title = "Hypergeometric Score by Label",
    x = "phyper",
    y = "Label",
    color = "Label"
  ) +
  theme_minimal(base_size = 16) + 
  theme(
    plot.title = element_text(size = 20, hjust = 0.5), 
    axis.title = element_text(size = 16),              
    axis.text = element_text(size = 14),                              
    legend.title = element_text(size = 16),            
    legend.text = element_text(size = 14),                            
    legend.position = "right"
  )

ggplot(reduced_data, aes(x = CSS, y = as.factor(Label), color = as.factor(Label))) +
  geom_jitter(width = 0, height = 0.1, size = 2, alpha = 0.7) +
  scale_color_manual(values = c("red", "limegreen")) +
  labs(
    title = "CSS by Label",
    x = "CSS",
    y = "Label",
    color = "Label"
  ) +
  theme_minimal(base_size = 16) + 
  theme(
    plot.title = element_text(size = 20, hjust = 0.5), 
    axis.title = element_text(size = 16),              
    axis.text = element_text(size = 14),                              
    legend.title = element_text(size = 16),            
    legend.text = element_text(size = 14),                            
    legend.position = "right"
  )

ggplot(reduced_data, aes(x = omega_025, y = as.factor(Label), color = as.factor(Label))) +
  geom_jitter(width = 0, height = 0.1, size = 2, alpha = 0.7) +
  scale_color_manual(values = c("red", "limegreen")) +
  labs(
    title = "Omega Shrinkage by Label",
    x = expression(Omega[025]),
    y = "Label",
    color = "Label"
  ) +
  theme_minimal(base_size = 16) + 
  theme(
    plot.title = element_text(size = 20, hjust = 0.5), 
    axis.title = element_text(size = 16),              
    axis.text = element_text(size = 14),                              
    legend.title = element_text(size = 16),            
    legend.text = element_text(size = 14),                            
    legend.position = "right"
  )

custom_pr_curve <- function(data, truth_col, score_col) {
  data <- data[order(data[[score_col]], decreasing = TRUE), ]
  
  is_positive <- data[[truth_col]] == 1
  
  cum_true_positives <- cumsum(is_positive)
  cum_false_positives <- cumsum(!is_positive)
  
  total_positives <- sum(is_positive)
  
  precision <- cum_true_positives / (cum_true_positives + cum_false_positives)
  recall <- cum_true_positives / total_positives
  
  thresholds <- data[[score_col]]
  
  pr_data <- data.frame(
    threshold = thresholds,
    precision = precision,
    recall = recall
  )
  
  return(pr_data)
}

pr_curves <- list()
pr_curves$RR <- custom_pr_curve(Metrics_computation_size2_1M_count, "Label", "RR")
pr_curves$phyper <- custom_pr_curve(Metrics_computation_size2_1M_count, "Label", "phyper")
pr_curves$CSS <- custom_pr_curve(Metrics_computation_size2_1M_count, "Label", "CSS")
pr_curves$omega_025 <- custom_pr_curve(Metrics_computation_size2_1M_count, "Label", "omega_025")
pr_curves_list <- lapply(seq_along(pr_curves), function(i) {
  pr_curves[[i]] %>%
    mutate(Curve = paste(names(pr_curves)[i]))
})
combined_pr_data <- bind_rows(pr_curves_list)

### For the PRR since it has only a single threshold
true_positives <- sum(Metrics_computation_size2_1M_count$Label == 1 & Metrics_computation_size2_1M_count$PRR == 1)
false_positives <- sum(Metrics_computation_size2_1M_count$Label == 0 & Metrics_computation_size2_1M_count$PRR == 1)
total_positives <- sum(Metrics_computation_size2_1M_count$Label == 1)
precision_single <- true_positives / (true_positives + false_positives)
recall_single <- true_positives / total_positives


single_pr_point <- data.frame(
  recall = recall_single,
  precision = precision_single,
  Curve = "PRR"
)

combined_pr_data_with_point <- bind_rows(
  combined_pr_data,
  single_pr_point
)

ggplot(combined_pr_data_with_point, aes(x = recall, y = precision, color = Curve, linetype = Curve)) +
  geom_line(data = combined_pr_data, linewidth = 1.1) +
  geom_point(data = single_pr_point, size = 6, shape = 17) +
  labs(title = "Precision-Recall Curves for Cocktail Score",
       x = "Recall",
       y = "Precision",
       color = "Score",
       linetype = "Score") +
  theme_minimal() +
  guides(
    color = guide_legend(keywidth = unit(1, "lines"), keyheight = unit(0.8, "lines")),
    linetype = guide_legend(keywidth = unit(1, "lines"), keyheight = unit(0.8, "lines"))
  )

```


Comparison of scores for cocktail of size-two. Each dot denotes a
cocktail, and its risk computed on Dataset $D_2$ (@sec-simu_data) is shown on the x-axis. Red dots represent
cocktails that do not induce adverse event (negative controls), green
ones represent cocktail inducing adverse event (positive controls). The
bottom-right corner shows the Precision-Recall curves for each score.
The perfect classification corresponds to the upper right corner. The
areas under the curves allow to compare different methods. RR, PRR, Omega shrinkage,
phyper and the CSS are defined in @sec-risk-definition.
:::

## Application to Dataset $D_1${#sec-appli_d1}

### Estimation of Risk Distribution

Risk distribution was estimated for size-two drug cocktails on Dataset $D_1$ (@sec-simu_data). Estimation of risk distribution for higher
cocktail sizes is possible but it is nearly impossible to compare it
to the true distribution as it is computationally prohibitive to obtain.
The distribution estimated by the MCMC algorithm, is compared to the
true risk distributions in @fig-distri_simu.

The left panels display the distributions of risk scores for both the
estimated (top-left) and true (bottom-left) risk values. Both
distributions share a similar shape, with the majority of risk scores
concentrated at low values, as $95\%$ of the scores fall below 11.
However, some differences can be observed, particularly in the tail of
the distribution, where cocktails of higher risks are under-represented
in the estimated distribution.

<!-- ![Comparison of estimated and true risk distributions for size-two drug cocktails. Left panels show comparison of risk distribution among size-two cocktails, right panels allows to compare probabilities and quantiles of both distributions](figures/simu_plt.svg){#fig-distri_simu}-->

::: {#fig-distri_simu}
```{r}
#| layout: [[50,50], [50,50]]
#| fig-show: hold
#| warning: false
#| message: false
#| echo: false

load("figures/datasets/plot_MCMC_simu.RData")

plot_estimated_distribution <- plot_frequency(
  aprox_distri_size_2$Filtered_score_distribution[2:length(aprox_distri_size_2$Filtered_score_distribution)],
  binwidth = .30, sqrt = FALSE, xlab = "H(C)"
) + 
  labs(title = "Estimated Distribution of Risk Among Size 2 Cocktails") +
  ylim(0, 0.35) + 
  xlim(0, 55) +
  theme(
    plot.title = element_text(size = 15, hjust = .5),
    axis.text.x = element_text(size = 12)
  )
plot_estimated_distribution

# true Size 2
plot_true_distribution <- plot_frequency(
  true_dist_size2_2_3med$Filtered_score_distribution[2:length(true_dist_size2_2_3med$Filtered_score_distribution)],
  binwidth = .30, sqrt = FALSE, xlab = "H(C)"
) + 
  labs(title = "True Distribution of Risk Among Size 2 Cocktails") +
  ylim(0, 0.35) + 
  xlim(0, 55) +
  theme(
    plot.title = element_text(size = 15, hjust = .5),
    axis.text = element_text(size = 12)
  )

# QQ-plot size 2
qq_plot_output(aprox_distri_size_2,
               true_dist_size2_2_3med, filtered = T) +
  labs(title = str_wrap("QQ Plot of Estimated vs True Distribution of Risk Among Size 2 Cocktail",50)) +
  theme(
    plot.title = element_text(size = 16, hjust = .5),
    axis.text = element_text(size = 13)
    )

plot_true_distribution
# PP-plot size 2

estimated_data <- histogramToDitribution(aprox_distri_size_2$Filtered_score_distribution[2:length(aprox_distri_size_2$Filtered_score_distribution)])
true_data <- histogramToDitribution(true_dist_size2_2_3med$Filtered_score_distribution[2:length(true_dist_size2_2_3med$Filtered_score_distribution)])

ecdf_estimated_data <- ecdf(estimated_data)
ecdf_true_data <- ecdf(true_data)

common_sorted_data <- sort(unique(c(estimated_data, true_data)))
cp_estimated_data <- ecdf_estimated_data(common_sorted_data)
cp_true_data <- ecdf_true_data(common_sorted_data)

pp_df <- data.frame(EmpiricalCP1 = cp_estimated_data, EmpiricalCP2 = cp_true_data)

pp_plot <- ggplot(pp_df, aes(x = EmpiricalCP1, y = EmpiricalCP2)) +
  geom_point(alpha = 0.5) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red") +
  theme_minimal() +
  labs(x = "Cumulative Probabilities of Estimated Distribution",
       y = "Cumulative Probabilities of True Distribution",
       title = str_wrap("PP Plot of Estimated vs True Distribution of Risk Among Size 2 Cocktail",50)) +
  theme(
    plot.title = element_text(size = 16, hjust = .5),
    axis.text = element_text(size = 13)
  )
pp_plot

```

Comparison of estimated and true risk distributions for size-two drug
cocktails on Dataset $D_1$. Left panels show comparison of risk distribution among
size-two cocktails, right panels allow comparison of probabilities and
quantiles of both distributions
:::

The right panels of @fig-distri_simu present a Probability-Probability
(PP) plot (top) and a Quantile-Quantile (QQ) plot (bottom), comparing
the quantiles and probabilities of the estimated and true risk
distributions. While the right panel of the figure demonstrates good
agreement at lower risks, where most of the data lie, deviations at
higher risk values suggest that the estimated distribution slightly
underrepresents the risk for more extreme values.

Results indicate that the method performs well in estimating risk scores
for the majority of cocktails, capturing the overall risk distribution
with reasonable accuracy. The slight underestimations which are present
for high-risk cocktails are not a problem since the interest of the
method is to assign p-values. P-values are still reliable as shown in
the PP-plot, @fig-distri_simu. Consequently, we can assign robust
empirical p-values to any size-two cocktails based on their risks.
Furthermore the calibration under the null hypothesis for the p-values
attributed by the algorithm is assessed in @sec-appB.

### Genetic Algorithm Output and Clustering

The genetic algorithm was applied to Dataset $D_1$ to identify
high-risk drug cocktails. Multiple runs of the algorithm were conducted
using different hyperparameter sets (population size, number of generations, parameter $\alpha$) to ensure the visit of different regions of the space of possible cocktails. The results were subsequently concatenated to create a comprehensive list of high-risk cocktails.

The genetic algorithm successfully identified nearly all high-risk
size-two and size-three drug cocktails in Dataset $D_1$
(@fig-cluster_simu). For size-two cocktails, the algorithm consistently
found the exact high-risk combinations. However, for size-three
cocktails, the algorithm sometimes identified cocktails that were very
close to the true high-risk combinations, missing only one drug from the
correct cocktail in a few cases (oftentimes, choosing parent nodes
instead of the actual nodes).

To streamline the analysis of the large set of results, significant
cocktails were filtered using the empirical p-value by setting a
threshold of $5\%$. Clustering techniques were then applied to group
similar cocktails together. As discussed in @sec-clust, the UMAP
algorithm has been used for dimensionality reduction, followed by the
DBSCAN clustering method. This post-processing step allows to reduce
redundancy by grouping cocktails that differed only slightly, such as by
substituting a drug for another within the same pharmacological family.
Such cocktails would have similar medical interpretations.

<!-- ![Clustering of high-risk drug cocktails identified by the genetic algorithm on the simulated dataset. Each dot represents a drug cocktail, true cocktail solutions are the centers of the squares](figures/clustering_simu.svg){#fig-cluster_simu} -->


```{r}
#| echo: false
#| message: false
#| warning: false
#| results: "hide"
#| out-width: "90%"
#| label: fig-cluster_simu
#| fig-cap: "Clustering of high-risk drug cocktails identified by the genetic algorithm on the simulated dataset. Each dot represents a drug cocktail, true cocktail solutions are the centers of the squares"


load("figures/datasets/clustering_genetic_algorithm.RData")

distribution_list <- list(aprox_distribution_size2_5Mstep_2_3med_t800,
                          aprox_distribution_size3_60Mstep_2_3med_t800)

file.copy("./figures/datasets/solutions_simu_no_p_value.csv", 
          "./figures/datasets/tmp_copy_solutions_simu_no_p_value.csv")

p_value_csv_file(distribution_list,
                 "figures/datasets/tmp_copy_solutions_simu_no_p_value.csv")


df_results_genetic_algorithm <- read.csv("figures/datasets/tmp_copy_solutions_simu_no_p_value.csv", sep=";")

file.remove("./figures/datasets/tmp_copy_solutions_simu_no_p_value.csv")

threshold <- .05

df_results_genetic_algorithm_signif <- df_results_genetic_algorithm %>%
  filter(!is.na(p_value), p_value <= threshold, n.patient.taking.C > 4)

df_results_genetic_algorithm_signif$cocktail_index <- 
  string_list_to_int_cocktails(ATC_name = ATC_Tree_UpperBound_2014$Name,
                               lines = df_results_genetic_algorithm_signif$Cocktail)

# We add the true signals of the simulated datasets in order to plot all cocktails
# after
empty_rows <- data.frame(matrix(NA,nrow=4,ncol=ncol(df_results_genetic_algorithm_signif)))
names(empty_rows) <- names(df_results_genetic_algorithm_signif)
df_results_genetic_algorithm_signif <- rbind(df_results_genetic_algorithm_signif, empty_rows)
df_results_genetic_algorithm_signif$cocktail_index[[(nrow(df_results_genetic_algorithm_signif)-3)]] <- c(47,1393,2199)
df_results_genetic_algorithm_signif$cocktail_index[[(nrow(df_results_genetic_algorithm_signif)-2)]] <- c(888,659,2338)
df_results_genetic_algorithm_signif$cocktail_index[[(nrow(df_results_genetic_algorithm_signif)-1)]] <- c(2590, 2740)
df_results_genetic_algorithm_signif$cocktail_index[[(nrow(df_results_genetic_algorithm_signif))]] <- c(919,3760)
df_results_genetic_algorithm_signif$true_cocktail <- 0
df_results_genetic_algorithm_signif$true_cocktail[(nrow(df_results_genetic_algorithm_signif)-3):(nrow(df_results_genetic_algorithm_signif))] <- 1


# Compute the dissimilarity for the clust
dissimilarity <- get_dissimilarity_from_cocktail_list(df_results_genetic_algorithm_signif$cocktail_index,
                                                      ATC_Tree_UpperBound_2014)

# Set the UMAP config
umap_config <- umap.defaults
umap_config$random_state <- 42
umap_config$input <- "dist"
umap_config$min_dist <- .001
umap_config$n_neighbors <- 13

umap_results <- umap(do.call(rbind, as.matrix(dissimilarity)), config = umap_config)

df_results_genetic_algorithm_signif$X <- umap_results$layout[,1]
df_results_genetic_algorithm_signif$Y <- umap_results$layout[,2]

# Set up the dbscan config
dbscan_results <- dbscan(umap_results$layout, eps= 2)
df_results_genetic_algorithm_signif$cluster <- dbscan_results$cluster

# Plot final results
ggplot(df_results_genetic_algorithm_signif, aes(
  x = X, 
  y = Y, 
  color = factor(cluster),
  shape = factor(true_cocktail),
  size  = factor(true_cocktail)
)) +
  geom_point(stroke = 1) + 
  scale_shape_manual(values = c("0" = 16, "1" = 0)) +  
  scale_size_manual(values = c("0" = .8, "1" = 4)) +
  labs(
    title = "Clustered Points with DBSCAN used on UMAP projection",
    shape = "True Signal",
    x = "UMAP1",
    y = "UMAP2",
    color = "DBSCAN Cluster"
  ) +
  guides(size = "none") +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 12, hjust = 0.5),
    axis.text = element_text(size = 10),
    axis.title = element_text(size = 11)
  )

```


@fig-cluster_simu illustrates the results of the clustering process after mapping the selected cocktails in the plane. Each point represents a drug cocktail, and the $7$ automatically determined clusters are represented by the colors.

The first conclusion is that the clusters are well separated in terms of the editing distance defined in @sec-diversity. The genetic algorithm thus identified seven regions of the cocktails space of high score that are different in terms of interpretation as their cocktails are far away between classes. On the other hand, cocktails in the same clusters are close, indicating a possible pharmacological interpretation of the cluster.

The second conclusion is that the algorithm effectively found the true solutions or at least very similar solutions. Indeed, the embedding of the true solutions yields the four squares, that clearly belong to four different clusters among the seven. The method therefore identifies and separates regions of the cocktail space containing the true solutions. That simulation thus enhances the confidence that investigating the high-score cocktails returned by the methods may allow to detect relevant phenomena.

## Application to the FAERS Spontaneous Reporting Data {#sec-appli-FAERS}

### Estimation of Risk Distribution

The risk estimation method was applied to the FAERS spontaneous reporting dataset presented in @sec-FAERS. @fig-qqplot_FAERS presents a
comparison between the estimated risk distribution and the true risk
distribution for size-two drug cocktails. The left panel of the figure
shows the distributions of risk scores, while the right panel presents
the QQ plot and the PP plot, comparing the quantiles and probabilities
of both distributions.

<!-- ![Comparison of estimated and true risk distributions for size-two drug cocktails on the FAERS dataset presented @sec-datasets. Left panels show comparison of risk distribution among size-two cocktails, right panels allows to compare probabilities and quantiles of both distributions](figures/FAERS_plt.svg){#fig-qqplot_FAERS} -->

::: {#fig-qqplot_FAERS}
```{r}
#| layout: [[50,50], [50,50]]
#| fig-show: hold
#| warning: false
#| message: false
#| echo: false
#| fig-width: 6
#| fig-height: 5
#| dpi: 300

load("figures/datasets/plot_MCMC_FAERS.RData")

plot_estimated_distribution <- plot_frequency(
  aprox_distribution_size2_5Mstep$Filtered_score_distribution[2:length(aprox_distribution_size2_5Mstep$Filtered_score_distribution)],
  binwidth = .40, sqrt = FALSE, xlab = "H(C)"
) + 
  labs(title = "Estimated Distribution of Risk Among Size 2 Cocktails") +
  ylim(0, 0.20) + 
  xlim(0, 80) +
  theme(
    plot.title = element_text(size = 15, hjust = .5),
    axis.text = element_text(size = 13)
  )
plot_estimated_distribution

# true Size 2
plot_true_distribution <- plot_frequency(
  true_distribution_myopathy_size2$Filtered_score_distribution[2:length(true_distribution_myopathy_size2$Filtered_score_distribution)],
  binwidth = .40, sqrt = FALSE, xlab = "H(C)"
) + 
  labs(title = "True Distribution of Risk Among Size 2 Cocktails") +
  ylim(0, 0.20) + 
  xlim(0, 80) +
  theme(
    plot.title = element_text(size = 15, hjust = .5),
    axis.text = element_text(size = 13)
  )


# QQ-plot size 2
qq_plot_output(aprox_distribution_size2_5Mstep,
               true_distribution_myopathy_size2, filtered = T) +
  labs(title = str_wrap("QQ Plot of Estimated vs True Distribution of Risk Among Size 2 Cocktail",50)) +
  theme(
    plot.title = element_text(size = 15, hjust = .5),
    axis.text = element_text(size = 13)
  )

plot_true_distribution

# PP-plot size 2

estimated_data <- histogramToDitribution(aprox_distribution_size2_5Mstep$Filtered_score_distribution[2:length(aprox_distribution_size2_5Mstep$Filtered_score_distribution)])
true_data <- histogramToDitribution(true_distribution_myopathy_size2$Filtered_score_distribution[2:length(true_distribution_myopathy_size2$Filtered_score_distribution)])

ecdf_estimated_data <- ecdf(estimated_data)
ecdf_true_data <- ecdf(true_data)

common_sorted_data <- sort(unique(c(estimated_data, true_data)))
cp_estimated_data <- ecdf_estimated_data(common_sorted_data)
cp_true_data <- ecdf_true_data(common_sorted_data)

pp_df <- data.frame(EmpiricalCP1 = cp_estimated_data, EmpiricalCP2 = cp_true_data)

pp_plot <- ggplot(pp_df, aes(x = EmpiricalCP1, y = EmpiricalCP2)) +
  geom_point(alpha = 0.5) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red") +
  theme_minimal() +
  labs(x = "Cumulative Probabilities of Estimated Distribution",
       y = "Cumulative Probabilities of True Distribution",
       title = str_wrap("PP Plot of Estimated vs True Distribution of Risk Among Size 2 Cocktail",50)) +
  theme(
    plot.title = element_text(size = 15, hjust = .5),
    axis.text = element_text(size = 13)
  )
pp_plot
```

Comparison of estimated and true risk distributions for size-two drug
cocktails on the FAERS dataset presented @sec-datasets. Left panels show
comparison of risk distribution among size-two cocktails, right panels
allow comparison probabilities and quantiles of both distributions. Left distributions are truncated to $80$ for visualization purposes. A few cocktails have a risk around $300$ as shown on the Y-axis of the QQ plot.
:::


The histogram reveals that the estimated distribution aligns well with
the true distribution for the majority of cocktails with lower risk
scores. However, deviations begin to emerge in the tail of the
distribution. Specifically, 15 of the riskiest cocktails in the true
distribution were not captured by the estimated distribution as we see
in the QQ plot. This explains the observed shift in the QQ plot at the
highest quantiles since higher risk cocktails have not been found by the
MCMC algorithm, highlighting the need of a complementary method like the
genetic algorithm in order to find riskiest cocktails.

Despite this slight deviation, the empirical p-values remain robust for
both lower and higher-risk cocktails as shown by the PP-plot in the
@fig-qqplot_FAERS.

### Genetic Algorithm Output and Clustering

The genetic algorithm was applied to the FAERS data focusing on the
myopathy AE by running $180$ parallel executions of the genetic
algorithm on varying population sizes (from $100$ to $1000$ cocktails
per generation). The whole procedure ran in less than $8$ hours on a
$24$-core server.

Cocktail sizes present in the merged final populations vary from $1$ to
$6$. The MCMC algorithm was run for each cocktail size in this range to
assign an empirical p-value to each solution. All solutions having a
p-value lower than $0.05$ were kept. No multiple testing correction was
made in order to avoid false negatives, that is disregarding interesting
cocktails, even if this may inflate the number of false positives. $682$
drug cocktails compose the final list. Applying Benjamini-Hochberg corrections to control the FDR at level $5\%$ and $10\%$ respectively kept the first $107$ and $564$ elements of the list.

The clustering procedure was applied to the $682$ cocktails, leading to $15$ clusters.
To validate the results, the first $150$ solutions were tagged with the families they correspond to, if any, without knowledge of the clustering's result. To do so, some drugs or drug families known to be linked to myopathy adverse events were considered, that is hypolipemic drugs,
Colchicine, corticosteroids, Ciclosporine, Beta blocking agents,
Fluoroquinolones, and anti-malaria drugs [@miernik2024drug;
@hall2011musculoskeletal; @valiyil2010drug]. 

The final result is a table of $682$ rows, one per selected cocktail,
indicating its composition, the number of patients taking it, the number
of patients taking it and facing myopathy, hypergeometric and RR score.
It also contains the cluster it belongs to and the tagged families. The
entire table is available [on this
link](https://docs.google.com/spreadsheets/d/1TmbrExi26QNr2TmvltmEzyl_EQht3fIsrtsWrAqmRik/edit?gid=2108693123#gid=2108693123).
@tbl-cluster_table summarizes the cluster assignments for the top $150$
cocktails identified in the genetic algorithm. Among these, eleven
cocktails could not be assigned to any drug or drug family listed in the
table headers. Note that a cocktail can be associated with more than one
drug or drug family.

| Cluster | Hypolipemic Drugs | Colchicine | Steroids | Cyclosporine | Beta blocking agents | Domperidone | Fluoroquinolones |
|:-------:|:-------:|:-------:|:-------:|:-------:|:-------:|:-------:|:-------:|
| 1 | **31** | 0 | 3 | 0 | 0 | 0 | 4 |
| 2 | 0 | **11** | 0 | 4 | 0 | 0 | 0 |
| 3 | 7 | 0 | 0 | 0 | **19** | 0 | 0 |
| 4 | 0 | 0 | 0 | 0 | 0 | **50** | 0 |
| 5 | 0 | 0 | 0 | 0 | 0 | **19** | 0 |
| 6 | 0 | 0 | 0 | 0 | 0 | 0 | **4** |

: Summary of clustering applied to the identified solutions. Each row
corresponds to a cluster. The $150$ cocktails with the highest risk were
analyzed. Box $(i,j)$ in the table represents the number of cocktails in
cluster $i$ that include the drug or drug family $j$
{#tbl-cluster_table}

This table shows the method's capability to detect known ADRs from the FAERS dataset and highlights the ability of the clustering method to group cocktails with similar pharmacological interpretations.
Specifically, cluster $1$ corresponds to cocktails containing Hypolipemic Drugs, cluster $2$ to Colchicine and Cyclosporine, cluster
$3$ to beta blocking agents and cluster $6$ to Fluoroquinolones. These
clusters align with previously reported pharmacological associations
[@miernik2024drug]. Similarly, clusters $4$ and $5$ correspond to
Domperidone-containing cocktails which have been associated with adverse
effects, including potential cardiac complications, as reported by the
British Medicines and Healthcare products Regulatory Agency [@MHRA2014].

Interestingly, additional drug families reported in [@miernik2024drug],
such as anti-malarial drugs, were also identified but at later ranks.
For instance, anti-malarial drugs are primarily grouped in cluster $9$,
as shown in the complete solution table.

### Post-Treatment Interaction Analysis{#sec-resFirth}

The penalized logistic regression described in @sec-firth was applied to the strongest signals identified by the genetic algorithm to assess whether
the detected cocktails reflect true pharmacological interactions. For the cyclosporine–colchicine combination, the interaction coefficient remained positive and highly significant ($p = 1.02 × 10^{−6}$) after accounting for individual drug effects, corroborating existing medical literature indicating a potentiation of muscle toxicity [@ducloux1997colchicine].

Conversely, the analysis of the identified size-four cocktail (metformin, prasugrel, bisoprolol, simvastatin) revealed that the risk signal was almost entirely explained by a size-three sub-cocktail (metformin, bisoprolol, simvastatin). In this instance, the addition of prasugrel did not yield a significant additional interaction effect, illustrating the importance of the post-treatment step in identifying the core of a detected signal.


# Conclusion

As co-medication becomes increasingly common, there is a growing need
for methods capable of detecting signals of harmful drug combinations
from the available large databases for further assessment. The proposed
method addresses this need by identifying signals and assigning them a
p-value using a hypergeometric disproportionality analysis measure.
Additionally, the method enables the identification of broader signals
within the ATC hierarchy by proposing "cocktails" of not only active
substances but also chemical, therapeutic, and anatomical families,
leveraging the hierarchical classification of active substances.

Application on synthetic datasets demonstrated that using the
hypergeometric score reduces the false positives from cocktails taken by
a small number of patients, enhancing the robustness of the measure. The
results on these datasets of our MCMC algorithm to estimate the
distributions of cocktail risks were encouraging, as the estimated
distributions closely aligned with the true distributions, indicating
reliable p-value assignment. Furthermore, the genetic algorithm
effectively identified the majority of the harmful cocktails, with a
high success rate, highlighting its efficiency in navigating the large
solution space.

Applying this method to previous FAERS data for the myopathy adverse
event yielded promising results. A literature review confirmed the
intersection between the identified signals and drugs known to have a
higher likelihood of causing myopathy, demonstrating the effectiveness
of the proposed methodology. Results also indicate that certain drug
combinations are more strongly associated with myopathy than individual
medications. Notably, the cyclosporine/colchicine combination exhibits a
higher hypergeometric score (73.1 vs. 63.4) and RR (879.4 vs. 57.1)
compared to colchicine alone, reinforcing the importance of analyzing
drug interactions in pharmacovigilance. This combination is known to be
more likely to induce myopathy [@ducloux1997colchicine].

Furthermore, our approach identified a size-four drug cocktail
(metformin, prasugrel, bisoprolol, simvastatin) associated with an
increased risk signal (hypergeometric score $= 72.2$, RR = $3060.6$).
This combination was observed in nine patients, all of whom experienced
the adverse event. While this finding demonstrates the feasibility of
detecting higher-order drug interactions, we emphasize that no clinical
validation is currently available for this specific combination. This
underscores both the potential of the method in identifying complex drug
interactions and the need for further validation through complementary
studies.

As demonstrated in @sec-resFirth, the post-treatment logistic regression step
provides clinical refinement by distinguishing true interactions from combined
individual effects.

This approach can also be extended to other settings where adverse
events are explored. For instance, the method could be applied using the
ICD diagnosis classification or the MedDRA system, both of which are
hierarchical classifications. Such an application would facilitate the
identification of symptoms associated with the consumption of drug
combinations.

The proposed method is implemented as an R package *emcAdr*, available
on GitHub and on the CRAN [@R-emcAdr]. A tool allowing the processing of
quarterly FAERS xml files to csv file directly usable by our method is
also available on
[GitHub](https://github.com/JulesBa-Git/FAERS-xml-parser).

For researchers and stakeholders it is crucial to remember that our method is hypothesis generating as are other signal detection methods in single medications. These methods aim to generate as few false negative results as possible but with the trade-off of more false positives. This means that any results should firstly be assessed for biological plausibility as well as validated in a more formal causal framework such as RCTs and target trial emulations. Furthermore, while spontaneous reporting systems have a key role in pharmacovigilance, they have known drawbacks as underlined in the introduction. They however provide important knowledge about the safety of drugs, and with our method also drug cocktails.

# Funding {.unnumbered}

This work was partially supported by the “PHC AURORA” programme (project
number: 49704QC), funded by the French Ministry for Europe and Foreign
Affairs, the French Ministry for Higher Education and Research and the
Norwegian Council for Research.

# Appendices {.appendix}

## Appendix A : Distance Pseudo-code Algorithm {#sec-appA .appendix}

``` pseudocode
#| label: algo-distance
#| html-indent-size: "1.2em"
#| html-comment-delimiter: "//"
#| html-line-number: true
#| html-line-number-punc: ":"
#| html-no-end: false
#| pdf-placement: "htb!"
#| pdf-line-number: true

\begin{algorithm}[H]
\caption{CalculateDistance}
\begin{algorithmic}[1]
\Require Matrix $M$, Lists $idxC_1$, $idxC_2$
\Ensure Cost of the distance calculation
\State $height \gets 0$
\State $cost \gets 0$
\While{$idxC_1 \neq \varnothing$ \textbf{and} $idxC_2 \neq \varnothing$}
  \ForAll{$iC_1 \in idxC_1$}
    \ForAll{$iC_2 \in idxC_2$}
      \If{$M[iC_1][height] = M[iC_2][height]$}
        \State $cost \gets cost + height - \min\!\Big(M[iC_1].\mathrm{count}( M[iC_1][0]),\ \!M[iC_2].\mathrm{count}(M[iC_2][0])\Big) + 1$
        \State \textbf{remove} $iC_1$ \textbf{from} $idxC_1$
        \State \textbf{remove} $iC_2$ \textbf{from} $idxC_2$
        \State \textbf{break}
      \EndIf
    \EndFor
  \EndFor
  \State $height \gets height + 1$
\EndWhile
\State $insertion\_cost \gets \mathrm{ATC\_HEIGHT} / 2$
\State $cost \gets cost + (|idxC_1| + |idxC_2|) \times insertion\_cost$
\State \Return $cost$
\end{algorithmic}
\end{algorithm}
```

@algo-distance computes the distance between two cocktails, $C_1$ and
$C_2$, based on three inputs.

The first input is an integer matrix $M$. This matrix represents the
nodes of $C_1$ and $C_2$, as well as their corresponding parent nodes.
The matrix has dimensions $(PN, \text{ATC\_HEIGHT})$, where $P$ and $N$
denote the sizes of $C_1$ and $C_2$, respectively. For example, consider
the simplified tree shown in Figure 1 of the article. In this tree,
$C_1$ could be the cocktail containing only the node
$\begin{bmatrix} 3 \end{bmatrix}$, and $C_2$ could be the cocktail
containing only the node $\begin{bmatrix} 2 \end{bmatrix}$ ($P = 1$ and
$N = 1$).

The corresponding matrix $M$ would be: 
$$
M = \begin{bmatrix}
    3 & 1 & 0 \\
    2 & 2 & 0
\end{bmatrix}
$$

The second input consists of $idxC_1$ and $idxC_2$, which represent the
indices of the rows in $M$ that contain the nodes of $C_1$ and $C_2$,
respectively.

## Appendix B : Calibration Under the Null {#sec-appB .appendix}

To assess the validity of the approximation of the p-value, we plotted the empirical
distribution of p-values for cocktails presumed not to increase the
adverse-event rate on the Dataset $D_1$. The theoretical p-values  following a Uniform(0, 1) distribution, the approximated ones should exhibit a distribution close to uniform if the approximation is valid.
@fig-calibration shows a histogram of these p-values, which appears
approximately flat.

![Histogram of p-values for cocktails presumed to satisfy the null
hypothesis on Dataset $D_1$ (no elevated adverse-event
risk)](figures/accuracy_under_the_null.svg){#fig-calibration}

## Appendix C : FAERS Clustering {.appendix}

@fig-cluster_FAERS shows the clustering applied to the results of the
genetic algorithm run on the FAERS datasets.

<!-- ![Clustering of High-Risk Drug Cocktails Identified by the Genetic Algorithm on the FAERS dataset. This Figure allows to see proximity between multiple clusters and better understand why some cluster like the $4$ and $5$ exhibits similarities (Domperidone related cocktails).](figures/clustering_FAERS_Myopathy.svg){#fig-cluster_FAERS}-->


```{r}
#| echo: false
#| message: false
#| warning: false
#| results: "hide"
#| out-width: "90%"
#| label: fig-cluster_FAERS
#| fig-cap: "Clustering of High-Risk Drug Cocktails Identified by the Genetic Algorithm on the FAERS dataset. This Figure allows to see proximity between multiple clusters and better understand why some cluster exhibits similarities."

load("figures/datasets/plot_genetic_FAERS.RData")

distribution_list <- list(drugs_distribution_myopathy,
                          aprox_distribution_size2_5Mstep,
                          aprox_distribution_size3_Myopathy_13Mstep,
                          aprox_distribution_size4_20Mstep,
                          aprox_distribution_size5_25Mstep,
                          aprox_distribution_size6_30Mstep)

file.copy("./figures/datasets/solutions_myopathy.csv", 
          "./figures/datasets/tmp_copy_solutions_myopathy.csv")

p_value_csv_file(distribution_list,
                 "figures/datasets/tmp_copy_solutions_myopathy.csv")


df_results_genetic_algorithm <- read.csv("figures/datasets/tmp_copy_solutions_myopathy.csv", sep=";")

file.remove("./figures/datasets/tmp_copy_solutions_myopathy.csv")

threshold <- .05

df_results_genetic_algorithm_signif <- df_results_genetic_algorithm %>%
  filter(!is.na(p_value), p_value <= threshold, n.patient.taking.C > 4)

df_results_genetic_algorithm_signif$cocktail_index <- 
  string_list_to_int_cocktails(ATC_name = ATC_Tree_UpperBound_2024$Name,
                               lines = df_results_genetic_algorithm_signif$Cocktail)


# Compute the dissimilarity for the clust
dissimilarity <- get_dissimilarity_from_cocktail_list(df_results_genetic_algorithm_signif$cocktail_index,
                                                      ATC_Tree_UpperBound_2024)

# Set the UMAP config
umap_config <- umap.defaults
umap_config$random_state <- 42
umap_config$input <- "dist"
umap_config$min_dist <- .01 
umap_config$n_neighbors <- 8

umap_results <- umap(do.call(rbind, as.matrix(dissimilarity)), config = umap_config)

df_results_genetic_algorithm_signif$X <- umap_results$layout[,1]
df_results_genetic_algorithm_signif$Y <- umap_results$layout[,2]

# Set up the dbscan config
dbscan_results <- dbscan(umap_results$layout, eps= 1.2)
df_results_genetic_algorithm_signif$cluster <- dbscan_results$cluster

# Plot final results
ggplot(df_results_genetic_algorithm_signif, aes(
  x = X, 
  y = Y, 
  color = factor(cluster)
)) +
  geom_point(size = .8) + 
  labs(
    title = "Clustered Points with DBSCAN used on UMAP projection",
    x = "UMAP1",
    y = "UMAP2",
    color = "DBSCAN Cluster"
  ) +
  guides(size = "none") +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 12, hjust = 0.5), 
    axis.text = element_text(size = 9),
    axis.title = element_text(size = 10),
    legend.text = element_text(size = 8),
    legend.key.height= unit(0.35, 'cm')
  )

```

## Appendix D : Algorithmic Complexity {.appendix}

The total time complexity of the high-risk cocktail identification is given by:
$$\mathcal{O}(G \cdot P \cdot T_{eval})$$
Where $G$ represents the number of generations (or MCMC iterations) and $P$ represents the population size of the GA (this equals one for the MCMC algorithm). The evaluation time for a single cocktail, $T_{eval}$, is defined by the matching logic required to count occurrences across the database:
$$T_{eval} = \mathcal{O}(N \cdot K \cdot R)$$

- $N$ (Number of Reports): The algorithm iterates once through the database to compute the hypergeometric score for a given cocktail. The runtime scales linearly with the number of patients.

- $K$ (Cocktail Size): The number of drugs in the candidate cocktail. In this study, $K$ is typically small since there is a limit to the number of drugs a patient takes.

- $R$ (Drugs per Report): The number of medications listed in a single patient report. On our data the mean size of drug cocktails per patient is approximately $2.2$. Because $K$ and $R$ are bounded, the complexity simplifies to $O(N)$ per candidate evaluation.

## Appendix E: Illustrative Example of Package Usage {.appendix}

This appendix demonstrates the end-to-end usage of the `emcAdr`
package on a small simulated dataset that compiles in under a
minute. It is intended as a hands-on illustration of the API: full
reproduction of the results in @sec-appli_d1 and
@sec-appli-FAERS requires the
longer-running scripts available in the article's GitHub repository. An accompanying `README.md` file in the article's GitHub repository documents the full reproduction workflow.

The toy dataset follows the same simulation design as Dataset $D_1$ (@sec-simu_data), at a reduced scale and with deliberately inflated adverse-event probabilities so that the signals remain detectable in a small sample. Patients taking only one of the two component drugs are added as decoys to break the spurious equivalence between the size-two cocktail and either of its components in a small dataset.

### E.1 Setup and Toy Dataset Generation

```{r}
#| label: setup-toy
#| message: false
#| warning: false

library(emcAdr)
library(dplyr)
library(purrr)
library(tidyr)
library(ggplot2)
library(umap)
# we load data in order to have the ATC_Tree_UpperBound_2014
load("./figures/datasets/simulated_dataset_2_3_med.RData")
set.seed(42)

# ATC tree leaves (active substances) -- 0-indexed for the C++ backend
leaves <- which(ATC_Tree_UpperBound_2014$ATC_length == 7) - 1L

# Two size-two cocktails solution, each defined by two drug-family ranges
solutions <- list(
  C1 = list(d1_range = 48:53,    d2_range = 1394:1396, ae_prob = 1/5),
  C2 = list(d1_range = 889:892,  d2_range = 660:661,   ae_prob = 1/7)
)

N           <- 5000
prop_full   <- 0.02    # patients taking BOTH drugs of a cocktail
prop_decoy  <- 0.03    # patients taking ONLY ONE drug of a cocktail
ae_baseline <- 1/300

# Decoy mechanism: for each cocktail we add patients who take ONLY one
# of the two component drugs, with no elevated AE risk. Without these
# decoys, in such a small dataset, every taker of d1 would also take d2,
# so H({d1}) would match H({d1, d2}) and the algorithm would surface size-1
# solutions "equivalent" to the size-2 cocktail solutions. Decoys break that
# perfect correlation: a sizeable share of d1-takers are now AE-free.
# This mirrors the "2 out of 3" decoys used for D1 in Section 2.6.1.
group_spec <- imap(solutions, \(cock, cname) list(
  full     = list(p = prop_full,  ae_p = cock$ae_prob,
                  draw = \() c(sample(cock$d1_range, 1),
                               sample(cock$d2_range, 1))),
  decoy_d1 = list(p = prop_decoy, ae_p = ae_baseline,
                  draw = \() sample(cock$d1_range, 1)),
  decoy_d2 = list(p = prop_decoy, ae_p = ae_baseline,
                  draw = \() sample(cock$d2_range, 1))
)) |>
  list_flatten(name_spec = "{outer}_{inner}")

# The remaining patients receive a random Poisson-sized cocktail
group_spec$random <- list(
  p    = 1 - sum(map_dbl(group_spec, "p")),
  ae_p = ae_baseline,
  draw = \() sample(leaves, 1 + rpois(1, 4), replace = FALSE)
)
stopifnot(group_spec$random$p > 0)

# Sample a group label for each patient, then realize their data
group <- sample(names(group_spec), N, replace = TRUE,
                prob = map_dbl(group_spec, "p"))

# Note : columns of observations must have those names
toy_data <- tibble(group = group) |>
  mutate(
    patientATC = map(group, \(g) group_spec[[g]]$draw()),
    patientADR = map_lgl(group, \(g) runif(1) <= group_spec[[g]]$ae_p)
  ) |>
  select(patientATC, patientADR)

# Patient counts and AE counts by group
tibble(group = group, ADR = toy_data$patientADR) |>
  count(group, ADR) |>
  pivot_wider(names_from = ADR, values_from = n,
              names_prefix = "AE_", values_fill = 0L)
```

The dataset contains `r N` patients, of whom approximately
`r 100 * prop_full`% take the full cocktail $C_1$, `r 100 * prop_full`%
the full cocktail $C_2$, and `r 100 * prop_decoy`% take a single drug
from each cocktail as a decoy. The remaining patients receive a
random combination of active substances drawn from the ATC tree leaves.

C1_full patients show a $20$% AE rate vs $0.4$% in the random group. Decoys show baseline AE rates as designed.

Please note that the column of the patient dataframe has to be `PatientATC` and 
`PatientADR`.

### E.2 Evaluating Cocktail Risk via the Hypergeometric Score

We start by computing the hypergeometric score $H(C)$
(@sec-risk-definition) for a solutions cocktail and for a random
combination, to confirm that $C_1$ is clearly separated on the score
scale.

```{r}
#| label: hyper-score

# Pick the C1 family and a random size-2 cocktail
solution_C1 <- c(47, 1393)

random_C   <- sample(leaves, 2)

compute_hypergeom_on_list(list(solution_C1, random_C), 
                     ATCtree      = ATC_Tree_UpperBound_2014,
                     observations = toy_data)

```

In datasets of this scale, the score function is sparse: most randomly drawn size-2 pairs are taken by zero or one patient. A random pair therefore has near $100$% probability of scoring exactly zero, as illustrated by random_C above. The solution family $C_1$ scores $H(C_1) \approx 38$, but whether this magnitude is extreme among cocktails actually present in the data is the question answered by the next section.

### E.3 Estimating the Null Distribution via MCMC

A short Metropolis-Hastings run (@sec-MCMC) approximates the
distribution of $H(C)$ for size-two cocktails. The full-scale
analysis uses substantially longer chains.

```{r}
#| label: mcmc-null

mcmc_chain <- DistributionApproximation(
  epochs        = 200000,
  ATCtree       = ATC_Tree_UpperBound_2014,
  observations  = toy_data,
  temperature   = 25,
  Smax          = 2, # Size of cocktail
  num_thread    = 4
)

plot_frequency(
  mcmc_chain$ScoreDistribution[-1],
  binwidth = .3, xlab = "H"
)

# Convert the histogram-style MCMC output to a flat score vector.
# Index 1 corresponds to score = 0 cocktails, we drop it since we work only on
# non zero score. 
mcmc_null <- histogramToDitribution(mcmc_chain$ScoreDistribution[-1])

# Empirical p-value of a solution cocktail under the estimated null
solution_score <- compute_hypergeom_on_list(
  list(solution_C1), 
  ATC_Tree_UpperBound_2014,
  toy_data
)

cat("p-value of  solution C1 (", solution_C1, ") :",
    mean(mcmc_null >= solution_score))
```
The bulk of the distribution sits near zero. The visible spike around $H(C) \approx 4-5$  corresponds to a small-sample artifact, cocktails taken by only one patient where the patient happens to experience the AE. With only $\sim 45$ AE cases over $5000$ patients, such configurations are common enough to form a visible noise peak. Furthermore, with only $5000$ data points, each cocktail tend to be taken by few patient. The solution families generate $H(C)$ values in the $18$-$40$ range, above this peak, which is why the empirical p-value of $C_1$ comes out at $\sim 2.5 \times 10^{-3}$. 

For size-two cocktails, computing the *exact* score distribution is tractable and offered by the package via `trueDistributionSizeTwoCocktail()`. This allows assessing the MCMC approximation directly (see also @fig-distri_simu in the body):

```{r}
#| label: mcmc-vs-true
#| eval: false

true_dist <- trueDistributionSizeTwoCocktail(
  ATCtree      = ATC_Tree_UpperBound_2014,
  observations = toy_data,
  beta         = 4,
  num_thread = 4
)

# Compare via the package's qq-plot helper
qq_plot_output(mcmc_chain, true_dist)
```

The PP/QQ-plot comparison would confirm that the MCMC tracks the null faithfully on the regions of $H$ that actually occur in the data.

### E.4 Identifying High-Risk Cocktails with the Genetic Algorithm

To mirror the multi-run aggregation workflow of @sec-appli_d1, we run the genetic algorithm five times with different seeds and pool the final populations. A single short run typically converges on one of the two solutions. Aggregation across runs increases the chance of recovering both. For full-scale runs on $D_1$, parallel execution scripts compatible with SLURM and similar workflow managers are available in the article's GitHub repository under `figures/scripts/`.

```{r}
#| label: ga-run

n_runs <- 5
ga_runs <- map(seq_len(n_runs), \(i) {
  set.seed(100 + i)
  GeneticAlgorithm(
    epochs        = 400,
    nbIndividuals = 200,
    ATCtree       = ATC_Tree_UpperBound_2014,
    observations  = toy_data,
    num_thread    = 4,
    alpha         = 1.6, 
    diversity     = TRUE,
    p_mutation    = .09,
    summary       = FALSE
  )$FinalPopulation |> as_tibble()
})

ga_top <- bind_rows(ga_runs) |>
  mutate(cocktails = lapply(cocktails, sort)) |>
  group_by(cocktails) |>
  summarise(score = max(score), .groups = "drop") |>
  filter(score > 0) |>
  arrange(desc(score)) |>
  mutate(cocktails_display = map_chr(cocktails, paste, collapse = ", "))

ga_top |> head(10)
```

The top outputs correspond to the solution families. Cocktails involving the parent ATC nodes for $C_1$ (nodes $0$, $45$, $47$ paired with $1391$, $1393$) score in the $36$-$40$ range. Those for $C_2$ (nodes $630$, $658$, $659$ paired with $888$, $863$) score around $18$-$19$. 

### E.5 Output Clustering

For didactic purposes we apply the same UMAP + DBSCAN pipeline as described in @sec-clust on the recovered cocktails. We expect only a handful of distinct signals at this scale.

```{r}
#| label: clustering

# We put solution C1 and C2 in order to compare them to the projection 
# of found cocktail in the plane.
ga_cocktails <- c(list(c(47, 1393), c(888, 659)), ga_top$cocktails)

# Compute the distance between cocktails
divergence <- get_dissimilarity_from_cocktail_list(
  ga_cocktails, ATC_Tree_UpperBound_2014, normalization = TRUE
)
divergence <- as.matrix(do.call(rbind, divergence))

umap_config = umap::umap.defaults
umap_config$n_neighbors <- 10
# Precise UMAP that the input is a dissimilarity matrix
umap_config$input <- "dist"
umap_config$min_dist <- .001

umap_results <- umap::umap(divergence, config = umap_config)

dbscan_results <- dbscan::dbscan(umap_results$layout, eps = 1)

ga_plot <- tibble(cocktails = ga_cocktails) |>
  mutate(
    UMAP1    = umap_results$layout[, 1],
    UMAP2    = umap_results$layout[, 2],
    cluster  = factor(dbscan_results$cluster),
    solution = ifelse(
      row_number() %in% 1:2, 1, 0 # First two points are the true signals
      )
  )

ggplot(ga_plot, aes(UMAP1, UMAP2, colour = cluster)) +
  geom_point(alpha = 0.8, size = 2,aes(shape=as.factor(solution))) +
  labs(colour = "Cluster", shape = "Solution") +
  theme_minimal()
```

The two triangles (true solutions) embed in clusters 1 and 2 alongside related GA cocktails, demonstrating that the dissimilarity-based clustering groups variants of the same underlying signal in a reasonable way.

### E.6 Post-Treatment with Firth's Penalized Logistic Regression

For the $C_1$ cocktails, we apply the post-treatment of
@sec-firth to test whether the signal reflects a true synergistic
interaction or merely the additive effect of its components. A function in 
`emcAdr` has been added in order to test a cocktail.

```{r}
#| label: firth-posttreat

firth_reg <- run_firth_regression(
  cocktail = solution_C1,
  upper_bound  = ATC_Tree_UpperBound_2014$upperBound,
  patient_data = toy_data, 
  adr_column   = "patientADR"
)

summary(firth_reg)
```

The p-value of $0.032$ on the interaction term confirms that this synergistic effect is statistically distinguishable, even with only 16 AE cases in the `C1_full` group.
It confirms that the genetic algorithm's discovery reflects a true drug–drug interaction rather than the marginal contribution of either family alone.

Genetic algorithm output also contains cocktail $1393$ which contributes 
significantly to the AE on his own as depicted in the summary.

# Code and Results {.appendix}

See the complete results of application of methodology on the FAERS
dataset [on this
link](https://docs.google.com/spreadsheets/d/1TmbrExi26QNr2TmvltmEzyl_EQht3fIsrtsWrAqmRik/edit?usp=sharing).

Code to reproduce the experiments conducted in this article can be found
here :

-   GitHub link of the data refinement code :
    [JulesBa-Git/FAERS-xml-parser](https://github.com/JulesBa-Git/FAERS-xml-parser).
-   GitHub link to the latest release of the R package of the complete
    methodology :
    [JulesBa-Git/emcAdr](https://github.com/JulesBa-Git/emcAdr).
-   The code of the methodology is also available on the
    [CRAN](https://cran.r-project.org/web/packages/emcAdr/index.html).
-   The code to obtain figures is available on the `figures/scripts`
    folder of this repository.

# References {.unnumbered}

::: {#refs}
:::

# Session information {.appendix .unnumbered}

```{r session-info}
sessionInfo()
```
